---
title: "Cook County Property Tax Model in R"
output: 
  github_document:
    keep_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### What is it?

A while back [Tanya Schlusser](http://tanyaschlusser.github.io/) published an [excellent blog
post](http://tanyaschlusser.github.io/posts/property-tax-cook-county/) along with a [very
accomplished Python
Notebook](https://github.com/tanyaschlusser/tanyaschlusser.github.io/blob/src/posts/property-tax-cook-county.ipynb)
modeling property taxes in her neighborhood. Turns out her neighborhood is next to mine, and Python
is close to R so figured I should examine / replicate / extend this.

This repo, package and write-up are my current crack at it. See the
[README.md](https://github.com/eddelbuettel/cook-county-tax-model) for some background. We bundled
the code in an R package (in the same repo) as packages are a good way to bundle up code and data.

Here, we use a pre-made csv file (or, rather, data.frame resulting from reading it). The steps from
getting initial Property Index Number (PIN) data and accumulation of per-property data given each
PIN are described at the repo.

### Data

```{r data}
library(ccptm)                          # load this package
suppressMessages({
    library(data.table)                 # data manipulation
    library(ggplot2)                    # plotting
    library(GGally)                     # additional pairs plot
    })
data(riverforest)  # our sample from River Forest, IL -- raw

data <- data.table(riverforest)

## shorten some column names
nm <- names(data)
setnames(data, nm, gsub("^Prop", "", nm))
setnames(data, nm, gsub("^Char", "", nm))
setnames(data, nm, gsub("^Info", "", nm))

## minimal transformation
data[, `:=`(MktValCurrYear       = as.numeric(gsub("[$,]", "", MktValCurrYear)),
            MktValPrevYear       = as.numeric(gsub("[$,]", "", MktValPrevYear)),
            AsdValTotalCertified = as.numeric(gsub("[$,]", "", AsdValTotalCertified)),
            AsdValTotalFirstPass = as.numeric(gsub("[$,]", "", AsdValTotalFirstPass)),
            SqFt                 = as.numeric(gsub("[,]", "", SqFt)),
            BldgSqFt             = as.numeric(gsub("[,]", "", BldgSqFt)),
            CentAir              = as.factor(ifelse(CentAir == "Yes", 1L, 0L)),
            Frpl                 = as.ordered(as.integer(Frpl)),
            Age                  = as.integer(Age),
            FullBaths            = as.ordered(as.integer(FullBaths)),
            HalfBaths            = as.ordered(as.integer(HalfBaths)),
            Basement             = as.factor(Basement),
            Garage               = as.factor(Garage),
            Attic                = as.factor(Attic),
            Use                  = as.factor(Use)
            )]
## re-order / re-level data
data[, `:=`(Basement  = stats::reorder(Basement, MktValCurrYear, FUN=median),
            Attic     = stats::reorder(Attic, MktValCurrYear, FUN=median))]
data[, `:=`(Basement  = stats::relevel(Basement, "None"),
            Attic     = stats::relevel(Attic, "None"))]
data[]
```

### Initial Plot

The market value and assed value columns are linearly dependent via a simple '10x'
multiplication. We will focus the `MktValCurrYear` column, either of the assessed value columns
could be used equally well.

We start with a simple histogram plus density estimate and added rug plot of data points (using
alpha blending to account for very high density density in the middle price range).m

```{r histplus}
## histogram + density + rug
ggplot(data, aes(x=MktValCurrYear)) +
    geom_histogram(aes(y=..density..), bins=40, color="darkgrey", fill="lightblue") +
    geom_density(color="darkblue") + geom_rug(alpha=0.1)
```

### Scatter Plot

Several additional variables are of interest. The following pairs plot show several of them. 

We split this into two plots to not scale each pairs-plot cell down by too much.

```{r scatter1}
## pairs
## Frpl, FullBaths, HalfBaths, MktValCurrYear) -- CentAir,
ggpairs(data[, .(MktValCurrYear, SqFt, BldgSqFt, Age)],
        diag = list(continuous="densityDiag"),
        upper = list(continuous = wrap("points", size=0.5, alpha=0.1),
                     discrete="facetbar"),
        lower = list(continuous = wrap("points", size=0.5, alpha=0.1),
                     combo = "facetdensity"))
```

```{r scatter2}
## pairs
## ,
ggpairs(data[HalfBaths<=2 & FullBaths<=4,   # suppresses some warnings from empty cells
             .(MktValCurrYear, Frpl, FullBaths, HalfBaths, CentAir)],
        mapping = aes(alpha = 0.001),
        diag = list(continuous="densityDiag"),
        upper = list(continuous = "points",
                     discrete="facetbar"),
        lower = list(continuous = "points",
                     combo = "facetdensity"))
```

### Linear Models

We fit a few first exploratory models. We omitted the summary for brevity here.

```{r fits1}
## minimal first fit, overall fairly weak
fit0 <- lm(log(MktValCurrYear) ~ log(SqFt) + log(BldgSqFt) + CentAir + Frpl + log(Age) +
              FullBaths + HalfBaths + Basement + Attic - 1, data)
#summary(fit0)

fit1 <- lm(log(MktValCurrYear) ~ log(SqFt) + log(BldgSqFt) + log(Age) + FullBaths + HalfBaths - 1, data)
#summary(fit1)

fit2 <- lm(log(MktValCurrYear) ~ log(SqFt) + log(BldgSqFt) + log(Age) - 1, data)
#summary(fit2)
```

We can also plot the fits:

```{r plotfits1}
ggFittedVsActual <- function(fitted, actual) {
    ll <- range(range(fitted), range(actual))
    ggplot(data.frame(x=fitted, y=actual)) +
        geom_abline(intercept=0, slope=1, color="darkgrey") + #, style="dotted")
        geom_point(aes(x=fitted,y=actual),alpha=0.1,color="mediumblue") +
        coord_fixed(xlim=ll, ylim=ll) +
        xlab("Model Predictions") + ylab("Actual Values")
}


fit4 <- lm(MktValCurrYear ~ SqFt + BldgSqFt + Age - 1, data)
summary(fit4)
ggFittedVsActual(predict(fit4), data[,MktValCurrYear])

fit5 <- lm(log(MktValCurrYear) ~ log(SqFt) + log(BldgSqFt) + Age - 1, data)
summary(fit5)
ggFittedVsActual(exp(predict(fit5)), data[,MktValCurrYear])
```


The basic log-linear model (see Tanya's write-up for more; this appears to be
prescribed by the County) appears to lack a little bit of additional
structure:


Adding a squared term for squared footage reduces this effect and appears to
control for underprediction for more highly-priced homes:


```{r plotfits2}
fit6 <- lm(log(MktValCurrYear) ~ log(SqFt) + I(log(BldgSqFt)^2) + log(BldgSqFt) + Age - 1, data)
summary(fit6)
ggFittedVsActual(exp(predict(fit6)), data[,MktValCurrYear])
```


### Additional Variables: Attic ?

#### Versus MarketValue

A slight trend towards higher values with living area attics:

```{r attic1}
attic <- data[, .(Attic, MktValCurrYear)]
ggplot(attic, aes(y=MktValCurrYear, x=Attic, color=Attic)) +
    geom_boxplot(notch=TRUE) + geom_jitter(position=position_jitter(0.25), cex=0.75, alpha=0.5) +
    coord_flip() + theme(legend.position = "none")
```


#### Residuals: Linear Model

```{r attic2}
attic2 <- data[, .(Attic, Resid=resid(fit4))]
ggplot(attic2, aes(y=Resid, x=Attic, color=Attic)) +
    geom_boxplot(notch=TRUE) + geom_jitter(position=position_jitter(0.25), cex=0.75, alpha=0.5) +
    coord_flip() + theme(legend.position = "none")
```

#### Residuals: Log-linear Model

```{r attic3}
attic3 <- data[, .(Attic, Resid=exp(resid(fit5)))]
ggplot(attic3, aes(y=Resid, x=Attic, color=Attic)) +
    geom_boxplot(notch=TRUE) + geom_jitter(position=position_jitter(0.25), cex=0.75, alpha=0.5) +
    coord_flip() + theme(legend.position = "none")
```

#### Improved Fits?

So lets reduces the five-valued factor variable into just two values:  with or without attic.

```{r fitsWithAttic}
data[, finishedAttic := grepl("Living", as.character(Attic))]
fit4a <- lm(MktValCurrYear ~ SqFt + BldgSqFt + Age + finishedAttic - 1, data)
summary(fit4a)

fit5a <- lm(log(MktValCurrYear) ~ log(SqFt) + log(BldgSqFt) + Age + finishedAttic - 1, data)
summary(fit5a)
```

It does not seem to add much value beyond what is already covered by square
footage and building square footage.


### Additional Variables: Basement?

#### Versus MarketValue

```{r bsmt1}
bsmt <- data[, .(Basement, MktValCurrYear)]
ggplot(bsmt, aes(y=MktValCurrYear, x=Basement, color=Basement)) +
    geom_boxplot(notch=TRUE) + geom_jitter(position=position_jitter(0.25), cex=0.75, alpha=0.5) +
    coord_flip() + theme(legend.position = "none")
```

#### Residuals: Linear Model

```{r bsmt2}
bsmt2 <- data[, .(Basement, Resid=resid(fit4))]
ggplot(bsmt2, aes(y=Resid, x=Basement, color=Basement)) +
    geom_boxplot(notch=TRUE) + geom_jitter(position=position_jitter(0.25), cex=0.75, alpha=0.5) +
    coord_flip() + theme(legend.position = "none")
```

#### Residuals: Log-linear Model

```{r bsmt3}
bsmt3 <- data[, .(Basement, Resid=exp(resid(fit5)))]
ggplot(bsmt3, aes(y=Resid, x=Basement, color=Basement)) +
    geom_boxplot(notch=TRUE) + geom_jitter(position=position_jitter(0.25), cex=0.75, alpha=0.5) +
    coord_flip() + theme(legend.position = "none")
```

The results appear counter-intuitive: higher value without a basement? Tanya
discusses this a little bit.  One angle also is that there are very few
property without a basement.  The negative coefficient can be seen in light
of the added contribution to the other variable: an added basement, when
holding square footage constant, actually _takes away_ from non-basement
living space which could explain the sign of the coefficient.  Overall once
again a variable which we may keep excluded.

### Summary

We present a simple R implementation of a model of per-property attributes explaining the
assessed value as a proxy for market value. It replicates the analysis in the [original
writeup](http://tanyaschlusser.github.io/posts/property-tax-cook-county/) by [Tanya
Schlusser](http://tanyaschlusser.github.io/) on new data obtained with additional data
fetching code included with our model.
